---
title: 'R markdown for ST2195 Programming for Data Science Coursework'
author: "Zwe Lin Htet (Freddy)"
date: "3/30/2022"
output: html_document
---

```{r }
library(DBI)
library(dplyr)
library(ggplot2)
library(mlr3)
library(mlr3learners)
library(mlr3pipelines)
library(mlr3tuning)
library(mlr3filters)
library(mlr3fselect)
library(tidyr)
library(xgboost)
library(ranger)
library(glmnet)
```

```{r }
#Loading datasets into dataframe
year2005 <- read.csv("2005.csv", header = TRUE)
year2006 <- read.csv("2006.csv", header = TRUE)
year2007 <- read.csv("2007.csv", header = TRUE)
plane <- read.csv("plane-data.csv", header = TRUE)
airports <- read.csv("airports.csv", header = TRUE)
carriers <- read.csv("carriers.csv", header = TRUE)
variables <- read.csv("variable-descriptions.csv", header = TRUE)
```

```{r }
# Combining the chosen 3 years into one dataframe
df <- rbind(year2005, year2006, year2007)
```

Question 1
When is the best time of day, day of the week, and time of year to fly to minimise delays?

Best Day of Week
```{r }
#Finding average departure delay
depdelay_day <- df %>% 
  group_by(DayOfWeek) %>%
  summarize(AVGdepdelay_day = mean(DepDelay, na.rm = TRUE))
```

```{r }
#Finding average arrival delay
arrdelay_day <- df %>% 
  group_by(DayOfWeek) %>%
  summarize(AVGarrdelay_day = mean(ArrDelay, na.rm = TRUE))
```

```{r }
#Combing average departure delay and arrival delay
delay_day <- inner_join(depdelay_day, arrdelay_day, by = "DayOfWeek")
```

```{r }
#Finding total delay
DelayDay <- delay_day %>%
  group_by(DayOfWeek) %>%
  summarize(AvgDelay = (AVGdepdelay_day + AVGarrdelay_day)/2)
```

```{r }
#Coverting DayOfWeek into days 
days <- c("Mon", "Tue", "Wed", "Thurs", "Fri", "Sat", "Sun")
DelayDay$DayOfWeek <- days[DelayDay$DayOfWeek]
```

```{r }
#Plotting the graph for visualization
ggplot(DelayDay, aes(x=factor(DayOfWeek, 
                              levels = c("Mon", "Tue", "Wed", 
                                         "Thurs", "Fri", "Sat", "Sun")),
                     y=AvgDelay, 
                     fill=DayOfWeek)) + geom_bar(stat="identity", width = 0.6) + xlab("Days of Week") + ylab("Avg Delay Time") + ggtitle("Average Delay of the flights by day of a week")

```

Best Month of Year
```{r }
#Find average depdelay by months
depdelay_month <- df %>% 
  group_by(Month) %>%
  summarize(AVGdepdelay_month = mean(DepDelay, na.rm = TRUE))
```

```{r }
#finding average arrdelay by months
arrdelay_month <- df %>% 
  group_by(Month) %>%
  summarize(AVGarrdelay_month = mean(ArrDelay, na.rm = TRUE))
```

```{r }
#combining average depdelay and average arrdelay into one dataframe
delay_month <- inner_join(depdelay_month, arrdelay_month, by = "Month")
```

```{r }
#finding average delay 
DelayMonth <- delay_month %>%
  group_by(Month) %>%
  summarize(AvgDelay = (AVGdepdelay_month + AVGarrdelay_month)/2)
```

```{r }
#coverting Months into months 
months <- c("Jan", "Feb", "Mar", "Apr", "May", "June", "July", "Aug", "Sep", "Oct", "Nov", "Dec")
DelayMonth$Month <- months[DelayMonth$Month]
```

```{r }
#Plotting the graph for visualisation 
ggplot(DelayMonth, aes(x=factor(Month, 
                                levels = c("Jan", "Feb", "Mar", "Apr", "May",
                                           "June", "July", "Aug", "Sep", "Oct", "Nov", "Dec")),
                       y=AvgDelay, 
                       fill = Month)) + geom_bar(stat="identity", width = 0.6) + xlab("Months") + ylab("Avg Delay Time") + ggtitle("Average Delays of the flights by months of year")
```

Best Time of Day
```{r }
#finding the nearest hours
df$DepTime <- df$DepTime/100
df$DepTime <- floor(df$DepTime) 
```

```{r }
#remove NA vlaues in DepTime variable
df_time <- df %>% drop_na(DepTime)
```

```{r }
# select the DepTime and ArrDelay variable from the dataframe to find the average arrival delay 
delay_time <- df_time %>%
  select("DepTime", "ArrDelay") %>%
  group_by(DepTime) %>%
  summarize(AvgArrDelay = mean(ArrDelay, na.rm = TRUE))
```

```{r }
#removing the irrelevant roaws
delay_time <- delay_time[0:24,]
```

```{r }
ggplot(delay_time, aes(x = DepTime, 
                       y = AvgArrDelay, 
                       fill = DepTime)) + geom_bar(stat="identity", 
                                                   width = 0.6) + xlab("Hours") + ylab("Avg Delay Time (mins)") + ggtitle('Average Delay of the flights by time of a day')
```

Question 2
Do older planes suffer more delays?
```{r }
#changing column name from plane dataframe
names(plane)[1] <- "TailNum"
```

```{r }
#removing the unnecessary columns
plane <- plane[-c(2,3,4,5,6,7,8)]
```

```{r }
#removing rows that contain NA values 
plane <- plane[35:5029,]
```

```{r }
#merging two tables df and plane by TailNum
plane_data <- merge(x = df, y = plane, by = "TailNum", all.x = TRUE )
```

```{r }
#filtering the rows where TailNum is not zero 
plane_filter <- plane_data %>% 
  filter(TailNum != 0)
```

```{r }
#removing the row where year is NA
plane_remove <- subset(plane_filter, is.na(plane_filter$year) == F)
```

```{r }
#finding average arrdelay by year
year_arrdelay <- plane_remove %>% 
  group_by(year) %>%
  summarize(AVGarrdelay = mean(ArrDelay, na.rm = TRUE))
```

```{r }
#choosing only the relevant rows 
year_arrdelay <- year_arrdelay[3:50,]
```

```{r }
#plotting the graph
plot(year_arrdelay$year, year_arrdelay$AVGarrdelay, type = 'l', xlab = 'Year', ylab = 'Avg Arrival Delay', main = 'Average Arrival Delay of the planes by Year')
```

```{r }
#finding average depdelay by year
year_depdelay <- plane_remove %>% 
  group_by(year) %>%
  summarize(AVGdepdelay = mean(DepDelay, na.rm = TRUE))
```

```{r }
year_depdelay <- year_depdelay[3:50,]
plot(year_depdelay$year, year_depdelay$AVGdepdelay, type = 'l', xlab = 'Year', ylab = 'Avg Departure Delay', main = 'Average Departure Delay of the planes by Year')
```

Question 3
How does the number of people flying between different locations change over time?
```{r }
#counting the numbers of UniqueCarrieer by month and year
airline_count <- df %>%
  group_by(Month, Year) %>%
  count(UniqueCarrier)
```

```{r }
#finding the sum of flights by month and year
flight_count <- airline_count %>%
  group_by(Month, Year) %>%
  summarize(Sum_flight = sum(n))
```

```{r }
#Converting months 
months <- c("Jan", "Feb", "Mar", "Apr", "May", "June", "July", "Aug", "Sep", "Oct", "Nov", "Dec")
flight_count$Month <- months[flight_count$Month]
```

```{r }
ggplot(flight_count, aes(x = factor(Month, 
                                    levels = c("Jan", "Feb", "Mar", "Apr", 
                                               "May", "June", "July", "Aug", "Sep", "Oct", "Nov", "Dec")),  y = Sum_flight, group = Year)) + geom_line(aes(colour = factor(Year))) + geom_point() + scale_color_discrete(name = "Year", 
                                                                                                                      labels = c("2005", "2006", "2007")) + labs(x= "Months",  y= "Number of Flights", title = "Number of Flights in 2005, 2006 and 2007 by months")
```

```{r }
#finding sum of distance of flights by month and year
sum_distance <- df %>%
  group_by(Month, Year) %>%
  summarize(Sum_Distance = sum(Distance))
```

```{r }
#converting months
months <- c("Jan", "Feb", "Mar", "Apr", "May", "June", "July", "Aug", "Sep", "Oct", "Nov", "Dec")
sum_distance$Month <- months[sum_distance$Month]
```

```{r }
#plotting graph for visualisation 
ggplot(sum_distance, aes(x = factor(Month, levels = c("Jan", "Feb", "Mar", "Apr", "May", "June", "July", "Aug", "Sep", "Oct", "Nov", "Dec")), y = Sum_Distance, group = Year)) + geom_line(aes(colour = factor(Year))) + geom_point() + scale_color_discrete(name = "Year", labels = c("2005", "2006", "2007")) + labs(x= "Months", y= "Total Distance", title = "Sum of Distance of travelled by Airlines in 2005, 2006 and 2007")
```

Question 4
Can you detect cascading failures as delays in one airport create delays in others?
```{r }
#filtering the values where depdelay is greater than zero
dep <- df %>% 
  filter(DepDelay > 0)
```

```{r }
#filtering the values where arrdelay is greater than zero
arr1 <- dep %>%
  filter(ArrDelay > 0)
```

```{r }
#filtering the values where arrdelay is less than zero
arr2 <- dep %>%
  filter(ArrDelay < 0)
```

```{r }
#rounding of the values and finding the percentage values
arrdelay <- (nrow(arr1)/nrow(dep))*100
arrdelay_round <- round(arrdelay, 2)
arrdelay_percent <- paste0(arrdelay_round, "%")
```

```{r }
#rounding of the values and finding the percentage values
no_arrdelay <- (nrow(arr2)/nrow(dep))*100
no_arrdelay_round <- round(no_arrdelay, 2)
no_arrdelay_percent <- paste0(no_arrdelay_round, "%")
```

```{r }
#putting the arrdelay and no_arrdelay into dataframe
failure <- data.frame(arrdelay_percent, no_arrdelay_percent)
failure
```

Question 5
Use the available variables to construct a model that predicts delays.
```{r }
#setting sample size from orignal dataframe
size <- floor(0.001 * nrow(df))
```
```{r }
#defining train set and test set
train_set <- sample(size, round(0.5 * size))
test_set <- setdiff(1:size, train_set)
```

```{r }
#selecting the variables that we want to use for creating a model
delays <- df %>% 
  mutate(Delay = case_when(ArrDelay > 0 ~ 1, 
                           TRUE ~ 0)) %>%
  select(Month, DayofMonth, CRSElapsedTime, ActualElapsedTime, DepDelay, Delay, 
         WeatherDelay, SecurityDelay, CarrierDelay, LateAircraftDelay, NASDelay)
```      

```{r }
#Set up the task
delays$Delay <- factor(delays$Delay)
task <- TaskClassif$new('delays', backend = delays, target = 'Delay')
measure <- msr('classif.ce')
task
```

Some of the variables are factor for which some methods do not support factor, so we need to convert them to numerical values. (The following will be used later.)
```{r }
fencoder <- po("encode", method = "treatment",
               affect_columns = selector_type("factor"))
ord_to_int <- po("colapply", applicator = as.integer,
                 affect_columns = selector_type("ordered"))
```

Some methods require tuning the hyperparameters. (The following will be used later.)
```{r }
tuner <- tnr('grid_search')
terminator <- trm('evals', n_evals = 2)
```

Extend the results with different classification models
```{r }
# Logistic regression
learner_lr <- lrn('classif.log_reg')
gc_lr <- po('imputemean') %>>% po(learner_lr)

glrn_lr <- GraphLearner$new(gc_lr)

glrn_lr$train(task, row_ids = train_set)
glrn_lr$predict(task, row_ids = test_set)$score()
```

```{r }
# Gradient boosting
learner_gb <- lrn("classif.xgboost")
gc_gb <- po('imputemean') %>>%
  fencoder %>>% ord_to_int %>>%
  po(learner_gb)

glrn_gb <- GraphLearner$new(gc_gb)

glrn_gb$train(task, row_ids = train_set)
glrn_gb$predict(task, row_ids = test_set)$score()
```

```{r }
# Classification trees
learner_tree <- lrn("classif.rpart")
gc_tree <- po('imputemean') %>>%
  po(learner_tree)

glrn_tree <- GraphLearner$new(gc_tree)

glrn_tree$train(task, row_ids = train_set)
glrn_tree$predict(task, row_ids = test_set)$score()
```

```{r }
# Random Forest
learner_rf <- lrn('classif.ranger') 
learner_rf$param_set$values <- list(min.node.size = 4)
gc_rf <- po('scale') %>>%
  po('imputemean') %>>%
  po(learner_rf)

glrn_rf <- GraphLearner$new(gc_rf)

tune_ntrees <- ParamSet$new (list(
  ParamInt$new('classif.ranger.num.trees', lower = 50, upper = 600)
))

at_rf <- AutoTuner$new (
  learner = glrn_rf,
  resampling = rsmp('cv', folds = 2),
  measure = measure,
  search_space = tune_ntrees,
  terminator = terminator,
  tuner = tuner
)

at_rf$train(task, row_ids = train_set)
at_rf$predict(task, row_ids = test_set)$score() 
```

```{r }
set.seed(1) # for reproducible results

# list of learners
lrn_list <- list(
  glrn_lr,
  glrn_gb,
  glrn_tree,
  at_rf
)

# set the benchmark design and run the comparisons
bm_design <- benchmark_grid(task = task, resamplings = rsmp('cv', folds = 2), learners = lrn_list)
bmr <- benchmark(bm_design, store_models = TRUE)

autoplot(bmr) + theme(axis.text.x = element_text(angle = 45, hjust = 1))

bmr$aggregate(measure)
```






